---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: yes
---


## Loading and preprocessing the data
```{r}
data <- read.csv("activity.csv", header = TRUE)
x <- as.Date(data$date, format= "%Y-%m-%d")
head(data)
```

## What is mean total number of steps taken per day?

#### Calculating the total number of steps taken per day (spd)
```{r}
spd <- aggregate(data$steps, by = list(data$date), FUN = sum)
names(spd) <- c("date", "steps")
head(spd)
```
#### Making a histogram of the total number of steps taken per day
```{r}
hist(spd$steps, xlab = "number of steps", main = "Histogram of the total number of steps taken each day", col = "tomato2")
```

#### Calculating the mean and median of the total number of steps taken each day
```{r}
mean(spd$steps, na.rm = TRUE)
median(spd$steps, na.rm = TRUE)
```

## What is the average daily activity pattern?
```{r}
aspi <- aggregate(data$steps, by = list(data$interval), FUN = mean, na.rm = TRUE, na.action = na.omit)
names(aspi) <- c("interval", "mean")
plot(aspi$interval, aspi$mean, type = "l", pch = 19, col = "steelblue", 
     xlab = "5-minute interval", 
     ylab = "Average number of steps",
     main = "Time series plot of the average number of steps taken")
```

#### The 5-minute interval that, on average,contains the maximum number of steps
```{r}
aspi[which.max(aspi$mean),]
```

## Imputing missing values
#### Calculating the total number of missing values in the dataset
```{r}
sum(!complete.cases(data))
```

#### Filling in all of the missing values in the dataset with the mean for that 5-minute interval (data2)
```{r}
data2 <- data
for (i in 1 : length(data2$steps)) {
      if (is.na(data2$steps[i]) == TRUE) {
        data2$steps[i] <- aspi$mean[which(aspi$interval == data2$interval[i])]
      }
}
head(data)
```

#### Making a histogram of the total number of steps taken each day after missing values are imputed
```{r}
spd2 <- aggregate(data2$steps, by = list(data2$date), FUN = sum)
names(spd2) <- c("date", "steps")
hist(spd2$steps, xlab = "number of steps", main = "Histogram of the total number of steps taken each day", col = "yellow4")
```

#### Calculating the mean and median of the total number of steps taken each day after missing values are imputed
```{r}
mean(spd2$steps, na.rm = TRUE)
median(spd2$steps, na.rm = TRUE)
```

## Are there differences in activity patterns between weekdays and weekends?
#### Creating a new factor variable with two levels- "weekday" and "weekend"
```{r}
data2$day <- weekdays(as.Date(data2$date))
for (i in 1 : length(data2$day)) {
      if (data2$day[i] == "Saturday" || data2$day[i] == "Sunday") {
            data2$day[i] <- "Weekend"
      } else {
           data2$day[i] <- "Weekday"
      }
}
```
#### Calculating average number of steps per interval across all weekdays and weekends
```{r}
library(plyr)
aspi2 <- ddply(data2, .(interval, day), summarize, mean = round(mean(steps), 2))
head(aspi2)
```
#### Making a panel plot of the 5-minute interval and the average number of steps taken for weekdays and weekends
```{r}
library(lattice)
xyplot(mean ~ interval | day, 
       data = aspi2,
       type = "l",
       xlab = "5-minute interval",
       ylab = "Average number of steps",
       layout = c(1,2))
```